Epistemic Semantics in Guarded String Models

### 2020-06-21 tex/draft.tex
Edits in Sections 1 an 2. 2 is not finished.

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These worlds are often assumed to be things with the same
nature and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  To build such models, spaces of worlds and
individuals are stipulated as small finite sets, and semantic values
of lexical items are constructed as functions or relations from these
small sets.  Such toy or idealized models are useful in research and
in teaching, in that it is possible to represent propositions finitely
and explicitly, and to calculate with them. The point of this paper is
to scale up toy or idealized models to countable sets of worlds, and
to constructive and computable modeling of epistemic alternatives for
agents.  We describe a certain systematic way of defining such models,
and illustrate how to apply them in natural language semantics.
The focus on epistemic semantics and clausal embedding. The fundamental move is to
identify possible worlds with strings primitive events, so that
propositions are sets of strings. An advantage in this is that it
allows for a mathematical description of an algebra of a propositions,
coupled with a computational representation using either lazy lists of
strings, or finite state machines that describe sets of strings.

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.\marginpar{2001 or 2003?}  The
models we propose are an epistemic extension of guarded string models
for KAT, where a unary operation interpreted as an existential
epistemic modality is included for each agent.  {\em Action models} in
dynamic epistemic semantics introduced the technique of constructing
epistemic models from primitive alternative relations on events, in
order to capture the epistemic consequences of perceptual events
\citep{baltag1999logic}. This is the basis for our construction of
epistemic alternative relations.  Literature on finite state methods
in linguistic semantics has used event strings and sets of event
strings to theorize about tense and aspect in natural language
semantics \citep{fernando2004finite,fernando2007observing,carlson2009tense} and to
express intensionality \cite{fernando2017intensions}.  Literature on
finite state intensiononal semantics has discussed how to do the
semantics of intensional complementation including indirect questions
in a setting where compositional semantics is expressed in a finite
state calculus \citep{rooth2017finite,collard2018finite}.  We adopt
this in our syntax-semantics interface for English.

@@

{\em The Concealed Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up (H).  The coin could be H (heads) or T (tails), and
neither agent knows which it is.  This initial situation is possible world $w_1$.
Two additional worlds $w_2$ and $w_2$ are
defined by sequencing events after the initial state, with events interpreted as in
\ex{coin1}. The truth values for English
sentences shown in \ex{coin3} are observed.

\begin{example}  \label{coin1}
\hspace{-4mm} \begin{tabular}[t]{ll}
  $a_1$  & Amy peeks at H, by tipping the cup.
  \\ & Bob sees she's peeking, but not what \\
  & she sees. \\
  $b_1$  & Bob peeks at H. \\ 
\end{tabular}
\end{example}

@@

\begin{example} \label{coin3}
\hspace{-9mm} \begin{tabular}[t]{cccl}
  $w_1$ & $w_2$  & $w_3$ & Sentence\\
  false & true & true& Amy knows it's H. \\
  false & false & true& Bob knows it's H. \\
  false & false & true& Bob knows Amy \\ &&& knows it's H. \\
  false & true & true& Bob knows Amy knows \\ &&& whether it's H or T. \\
\end{tabular}
\end{example}

@@

@@

In this discussion, a sequence such as $\bar{q}{p}$ can be viewed as a
valuation of the primitive propositions $q$ and $p$, which is used to
describe world state. The primitives are listed in fixed order, and
left unmarked (indicating true) or marked with the overbar (indicating
false).  Suppose that in the coin example, we had an additional
primitive stative proposition (or atomic test) $t$, interpreted as
tails.  Since a coin is heads or tails but not both, we want to allow
the valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.  This is enforced by a {\em state formula}, which is
a simply Boolean formula, in this case \exi{effect}{a}.  Where
{\sf{B}} is a set of atomic tests and $\rho$ is a state constraint
over {\sf{B}}, ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\rho}$ is the set of
valuations of {\sf{B}} that make formula $\rho$ true. Following
\cite{kozen2001automata}, we call the valuations
\emph{atoms}.

Formulas like the ones in \ex{event1} that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between valuations, as defined in Figure
\ref{effectfigure}.  The valuations they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Aly peeking at heads) be as
specified in \ex{effect}. Then the set of valuations determined by
$\varphi$ and the relation on valuations for the event $u$ are
$\mathcal A_{\mathsf B}^\varphi$ and
$\llbracket\zeta\rrbracket^\varphi$ as given at the bottom in \ex{6}.
Given a state formula $\rho$, the the relation characterizing pre- and
post-conditions of an event with effect formula $\zeta$ is defined to
be $\sem{\zeta}^{\sem{\rho}}$.

\begin{example}   \label{effect}
\begin{tabular}[t]{l}
State formula: $\varphi = h\bar{t} + \bar{h}t$ \\
Effect formula for $a_1$: $\zeta = h \mathbin{:} h$ \\
\( \mathcal{A}^\varphi_{\mathsf B} = \{h\bar{t},\bar{h}t\} \)  \\
\( \sem{\zeta}^\varphi = \{\pair{h\bar{t}}{h\bar{t}}\} \)  \\
\end{tabular}
\end{example}

\begin{figure}
  \[\begin{array}{lcl}
  \multicolumn 3 l {\textrm{state formulas}\hfill (a \in \mathsf{B}) \hfill \mbox{}}\\
  \varphi,\psi &::=& a \mid0 \mid 1 \mid\varphi{+}\psi \mid \varphi~\psi \mid \bar{\varphi} \\
  \multicolumn {3} l {\textrm{effect formulas}}\\
  \zeta,\eta & ::= & \varphi \mathbin{:} \sigma \mid \zeta{+}\eta
    \mid \zeta \mathbin{\&} \eta \mid \bar{\zeta}
  \end{array}\]

\[
\begin{array}{cll}
  \llbracket\psi_1\mathbin{:}\psi_2\rrbracket^\varphi & = &
  \mathcal A^{(\varphi~\psi_1)}_{\mathsf{B}}\times\mathcal A^{(\varphi~\psi_2)}_{\mathsf{B}}\\[.1em]
\llbracket\zeta{+}\eta\rrbracket^\varphi & = & \llbracket\zeta\rrbracket^\varphi\cup\llbracket\eta\rrbracket^\varphi\\
\llbracket\zeta{\&}\eta\rrbracket^\varphi & = & \llbracket\zeta\rrbracket^\varphi\cap\llbracket\eta\rrbracket^\varphi\\
\llbracket\bar{\zeta}\rrbracket^\varphi & = & V \times V \setminus \llbracket\zeta\rrbracket^\varphi\\
\end{array}  
\]

\caption{Syntax and semantics of state formulas and effect formulas. Effect formulas
  are evaluated relative to as set Boolean valuations $V$, and denote relations between
  Boolean valuations.}
\label{effectfigure}
\end{figure}

\section{Epistemic guarded string models}
Figure \ref{epikfigure} shows an Epik program that describes a
possible worlds model for two agents with information about one coin,
and events of the agents semi-privately looking at the coin. The line
beginning with {\tt state} lists the basic stative propositions. To
illustrate syntax, a second proposition $t$ (tails) is included.  The
line beginning with {\tt constraint} defines compatibilities among the
propositions: the coin is heads or tails and not both.  The lines
beginning with {\tt event} declare events, their preconditions, and
their effect on state, following the format in \ex{event1}.  Finally
the lines beginning with {\tt agent} define {\em event alternative}
relations for agents.  Each clause with an arrow has a single event
symbol on the left, and a disjunction of alternative events on the
right of the arrow. The interpretation of Amy's alternatives for $b_1$
(Bill peeks at heads), is that when $b_1$ happens, for Amy either
$b_1$ or $b_0$ (Bill peeks at tails) could be happening.

Kleene Algebra with Tests is an algebraic theory that is defined by
equations and inequalities, which a has model classes including
guarded string models, relational models, finite models, and matrix
models.  This paper focuses on defining a family of concrete guarded
string algebras, the elements of which are sets of guarded strings.
Definitions and notation mostly follow \cite{kozen2001automata}.
Additional syntax and semantics is included to model multi-agent
epistemic semantics.  Guarded strings over a finite alphabet {\sf{P}}
are like ordinary strings, but with truth assignments to a set $T$ of
primitive propositions (primitive tests) alternating with the symbols
from $E$.  In the algebra described by Figure \ref{epikfigure}, $E$ is
the set of events $\{a_1,a_0,b_1,b_0\}$, and in the elevator example,
$\{u,d\}$.  In the elevator example, $T$ is $\{p,q\}$, and in the coin
example it is $\{h,t\}$.  We write atoms as the boolean formulae that
uniquely specify the truth assignment. In the elevator example, we
have atoms $\bar q \bar p$, $\bar q p$, $q \bar p$, $q p$, for
notational simplicity we will write these as $\hat0$, $\hat1$,
$\hat2$, $\hat3$ respectivelyto indicate the floor each atom
represents. In the coin example, we get $h\bar t$ and $\bar h t$ (for
which we use the shorthand $H$ and $T$), as well as the bogus atoms
$\bar h \bar t$ and $ht$!  To preclude bogosity, the
\texttt{constraint} declaration in Figure~\ref{epikfigure} describes a
state formula $\varphi$ which denotes the set of \emph{feasible atoms}
$\mathcal A_{\mathsf B}^\varphi$.

Guarded strings are strings of events, alternating with such atoms,
and starting and ending with atoms. The length of a guarded string
$g$, written $|g|$ is the number of events in $g$, ignoring the atoms.
\ex{gs1} gives the encoding as guarded strings of the worlds in
\ex{elevator1} and \ex{coin1}.

%% \begin{example}   \label{gs1}
%% \begin{tabular}[t]{ll}
%% $v_1$ & $\bar p \bar q ~u~ \bar p q ~u~ q \bar p ~d~ \bar p q ~u~ q \bar p$ \\
%% $v_2$ & $\bar p \bar q ~u~ \bar p q ~u~ p\bar q ~u~ pq$ \\   
%% $w_1$ & $\bar h t$ \\
%% $w_2$ & $\bar h t ~a_0~ \bar h t$ \\  
%% $w_3$ & $\bar h t ~a_0~ \bar h t ~b_0~ \bar h t$ \\
%% \end{tabular}
%% \end{example}
\begin{example}   \label{gs1}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}[t]{llc}
     \multicolumn 2 c {\textit{Guarded String}} & \textit{Length}\\ \hline
$v_1$ & $\hat 0 ~u~ \hat 1 ~u~ \hat 2 ~d~ \hat 1 ~u~ \hat 2$ & $4$ \\
$v_2$ & $\hat 0 ~u~ \hat 1 ~u~ \hat 2 ~u~ \hat 3$ & $3$ \\   
$w_1$ & $H$ & $0$\\
$w_2$ & $H ~a_0~ H$ & $1$ \\  
$w_3$ & $T~a_0~ T ~b_0~ T$ & $3$\\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

The discussion of \ex{coin2} mentioned building worlds by incrementing worlds with events.
This is accomplished in guarded string models with fusion product, a partial operation
that combines two guarded strings, subject to the condition that the truth assignment
at the end of the the first argument is identical to the truth assigment at the
start of the second one.  \ex{fusion1} gives some examples.

\begin{example} \label{fusion1}
\hspace{-2mm}\(\hat 0 ~u~ \hat 1 ~u~ \hat 2 ~d~ \hat 1 \diamond \hat 1 ~u~ \hat 2 = \hat 0 ~u~ \hat 1 ~u~ \hat 2 ~d~ \hat 1 ~u~ \hat 2 \)

\hspace{-2mm}\( \hat 0 ~u~ \hat 1 ~u~ \hat 2 ~d~ \hat 1 \diamond \hat 2 ~u~ \hat 3 = \mbox{\em{undefined}} \)

\hspace{-2mm}\(H \diamond H ~a_0~ H = H ~a_0~ H \)

\hspace{-2mm}\(T \diamond T ~a_1~ T = \mbox{\em{undefined}} \)
\end{example}

Rather than individual guarded strings, elements of a guarded string
model for KAT are sets of guarded strings.  In the application, these
elements have the interpretation of propositions (sets of possible
worlds) and/or event types.  An event such as $u$ in the guarded
string model corresponds to the set of guarded strings where the bare
event is flanked by compatible atoms. In standard KATs, every possible
atom is allowed to surround every event, which leads to nonsensical
strings like $\hat 1 ~u~\hat 0$ in the denotation of $u$, indicating
that the elevator travelled up from floor 1 to floor 0 (a physical
impossibility).

Fortunately, Epik provides a syntax to specify the input-output
behavior via the \texttt{event} declarations. From these we construct
a function $\mathcal E$ from agents to relations on atoms that specify
the effects each action can have. Notably for every declaration
$\texttt{event}~e~\zeta$, we have $\mathcal E^\varphi(e) =
\llbracket\zeta\rrbracket^\varphi$. We can lift $\mathcal E^\varphi$
to $\hat{\mathcal E}$ which decorates the events with their permitted
atoms, that is $\hat{\mathcal E}^\varphi(e) = \{\alpha~e~\beta \mid
\alpha, \beta \in \mathcal E^\varphi (e) \}$. See the examples in
\ex{events1}.
% u (\bar{q}\bar{p}):(\bar{q}p) + (\bar{q}p):(q\bar{p}) + (q\bar{p}):(qp) 173
% d (\bar{q}p):(\bar{q}\bar{p}) + (q\bar{p}):(\bar{q}p) + (qp):(q\bar{p})

\begin{example} \label{events1}
  \renewcommand{\arraystretch}{1.1}
\(
\begin{array}[t]{l|c}
  e &  \hat{\mathcal E}^\varphi(e) \\ \hline
  u & \{ \hat 0 ~u~ \hat 1, \hat 1 ~u~ \hat 2,  \hat 2 ~u~ \hat 3\}\\
  d & \{ \hat 1 ~d~ \hat 0, \hat 2~d~\hat 1, \hat 3 ~d~ \hat 2\}\\
  a_1 & \{H ~a_1~ H\}\\
  a_0 & \{T ~a_0~ T\}\\
  b_1 & \{H ~b_1~ H \}\\
  b_0 & \{T ~b_0~ T \}
\end{array}
\)
  \renewcommand{\arraystretch}{1}

@@

It remains to define the Kripke relation on guarded strings from an
agent specification as in Figure \ref{epikfigure}. In Epik, an agent
specification pairs each bare event with a set of ($+$-separated) bare
events, and so determines a relation between bare events, call it
relation $R_a$ for an agent $a$.  We can lift this to a relation
$\hat{R_a}$ between guarded strings, as defined in \ex{Rhathat}. The
operation $(\mathop;)$ in the definition is KAT product, which enforces
matching of tests.

@@

\begin{example}   \label{Rhathat}
\hspace{-2mm}\(
\hat{R} \triangleq
\hspace{-2mm}\begin{array}[t]{l}
\{ \tuple{x,y} \mid \exists u_1 ... \exists u_n \exists v_1 ... \exists v_n. \\
\hspace{6mm} u_1\mathrel{R}v_1 \wedge ... \wedge u_n\mathrel{R}v_n \wedge\\
\hspace{6mm} x \in \llbracket u_1 ; ... ; u_n \rrbracket^{\varphi,{\mathcal{E}}}\hspace{1mm} \wedge \\
\hspace{6mm} y \in \llbracket v_1 ; ... ; v_n \rrbracket^{\varphi,{\mathcal{E}}} \hspace{2mm} \}
\end{array}
\)
\end{example}

>>>>

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These are often assumed to be things with the same
nature and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  To build such models, spaces of worlds and
individuals are stipulated as small finite sets, and semantic values
of lexical items are constructed as functions or relations from these
small sets.  Such toy or idealized models are useful in research and
in teaching, in that it is possible to represent propositions finitely
and explicitly, and to calculate with them. The point of this paper is
to scale up toy or idealized models to countable sets of worlds, and
to constructive and computable modeling of epistemic alternatives for
agents.  We describe a certain systematic way of defining such models,
and illustrate how to apply them in natural language semantics.
The focus is on epistemic semantics and clausal embedding. The fundamental move is to
identify possible worlds with strings of primitive events, so that
propositions are sets of strings. An advantage in this is that it
allows for a mathematical description of an algebra of a propositions,
coupled with a computational representation using either lazy lists of
strings, or finite state machines that describe sets of strings.

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.\marginpar{2001 or 2003?}  The
models we propose are an epistemic extension of guarded string models
for KAT, where a unary operation interpreted as an existential
epistemic modality is included for each agent.  {\em Action models} in
dynamic epistemic semantics introduced the technique of constructing
epistemic models from primitive alternative relations on events, in
order to capture the epistemic consequences of perceptual events
\citep{baltag1999logic}. This is the basis for our construction of
epistemic alternative relations.  Literature on finite state methods
in linguistic semantics has used event strings and sets of event
strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensionality \cite{fernando2017intensions}.  Literature
on finite state intensiononal semantics has investigated how to do the
semantics of intensional complementation, including indirect
questions, in a setting where compositional semantics is expressed in
a finite state calculus \citep{rooth2017finite,collard2018finite}.  We
adopt this in our syntax-semantics interface for English.

@@

{\em The Concealed Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up.  The coin could be heads or tails, and
neither agent knows which it is.  This initial situation is possible world $w_1$.
Two additional worlds $w_2$ and $w_2$ are
defined by sequencing events after the initial state, with events interpreted as in
\ex{coin1}. The truth values for English
sentences shown in \ex{coin3} are observed.

\begin{example}  \label{coin1}
\hspace{-4mm} \begin{tabular}[t]{ll}
  $a_1$  & Amy peeks at heads, by tipping the
  \\ & cup. Bob sees she's peeking, but not\\
  & what she sees. \\
  $b_1$  & Bob peeks at heads. \\ 
\end{tabular}
\end{example}

@@

\begin{example} \label{coin3}
\hspace{-9mm} \begin{tabular}[t]{cccl}
  $w_1$ & $w_2$  & $w_3$ & Sentence\\
  false & true & true& Amy knows that it's \\ &&& heads. \\
  false & false & true& Bob knows thats it's \\&&&heads. \\
  false & false & true& Bob knows Amy \\ &&& knows it's heads. \\
  false & true & true& Bob knows Amy \\ &&& knows  whether it's \\ &&& heads or tails. \\
\end{tabular}
\end{example}

@@

\renewcommand{\arraystretch}{1.0}

@@

\section{Epistemic guarded string models}

In the discussion at the end of the last section, a sequence such as
$\bar{q}{p}$ can be viewed as a valuation of primitive test
propositions, which is used to describe world state. The primitives
are listed in fixed order, and left unmarked (indicating true) or
marked with the overbar (indicating false).  Suppose that in the coin
example, we have an additional primitive stative proposition (or
atomic test) $t$, interpreted as tails.  Since a coin is heads or
tails but not both, we want to allow the valuations $h\bar{t}$ and
$\bar{h}t$, and disallow $ht$ and $\bar{h}\bar{t}$.  This is enforced
by a {\em state formula}, which is a simply Boolean formula, in this
case \exi{effect}{a}.  Where {\sf{B}} is a set of atomic tests and
$\phi$ is a state constraint over {\sf{B}},
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\phi}$ is the set of valuations of
{\sf{B}} that make formula $\phi$ true. The valuations are called
atoms, because they correspond to the atoms of Boolean algebra of
tests \citep{kozen2001automata}.

Formulas like the ones in \ex{event1} that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between atoms, as defined in Figure
\ref{effectfigure}.  The atoms they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Aly peeking at heads) be as
specified in \ex{effect}. Then ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$
and the relation on atoms for the event $a_1$ are as given at
the bottom in \ex{effect}.

\begin{example}   \label{effect}
\(\begin{array}[t]{ll}
{\sf{B}} & \{h,t\}\\
\mbox{state formula $\varphi$} & h\bar{t} + \bar{h}t \\
\mbox{effect formula $\zeta$ for $a_1$} &  h \mathbin{:} h \\
 \mathcal{A}^\varphi_{\mathsf B} & \{h\bar{t},\bar{h}t\}   \\
\sem{\zeta}^\varphi & \{\pair{h\bar{t}}{h\bar{t}}\}   \\
\end{array}\)
\end{example}

\begin{figure}
  \[\begin{array}{lcl}
  \multicolumn 3 l {\textrm{state formulas}\hfill (a \in \mathsf{B}) \hfill \mbox{}}\\
  \rho,\sigma,\varphi &::=& a \mid0 \mid 1 \mid\rho+\sigma \mid \rho\,\sigma \mid \bar{\rho} \\ \\
  \multicolumn {3} l {\textrm{effect formulas}}\\
  \zeta,\eta & ::= & \rho \mathbin{:} \sigma \mid \zeta+\eta
    \mid \zeta \mathbin{\&} \eta \mid \bar{\zeta}
  \end{array}\]

\[
\begin{array}{lll}
  \sem{\rho\mathbin{:}\sigma}^\varphi & = &
  {\mathcal A}^{\rho\varphi}_{\mathsf{B}}\times{\mathcal A}^{\sigma\varphi}_{\mathsf{B}}\\[.1em]
\sem{\zeta +\eta}^\varphi & = & \sem{\zeta}^\varphi\cup\sem{\eta}^\varphi\\
\sem{\zeta \mathbin{\&}  \eta}^\varphi & = & \sem{\zeta}^\varphi\cap\sem{\eta}^\varphi\\
\sem{\bar{\zeta}}^\varphi & = & {\mathcal A}^{\varphi}_{\mathsf{B}} \times {\mathcal A}^{\varphi}_{\mathsf{B}} \setminus \sem{\zeta}^\varphi\\
\end{array}  
\]

\vspace{-2mm}

\caption{Syntax of state formulas and syntax and semantics of effect formulas. Effect formulas
  denote relations between atoms. In a state formula, juxtaposition $\rho\,\sigma$ is conjunction.}
\label{effectfigure}
\end{figure}

\vspace{1mm}

Figure \ref{epikfigure} shows an Epik program that describes a
possible worlds model for two agents with information about one coin,
and events of the agents semi-privately looking at the coin. The line
beginning with {\tt state} enumerates ${\sf{B}}$. The
line beginning with {\tt constraint} gives the state formula.
The lines
beginning with {\tt event} declare events and their effect formulas.
Finally
the lines beginning with {\tt agent} define {\em event alternative}
relations for agents.  Each clause with an arrow has a single event
symbol on the left, and a disjunction of alternative events on the
right of the arrow. The interpretation of Amy's alternatives for $b_1$
(Bill peeks at heads), is that when $b_1$ happens, for Amy either
$b_1$ or $b_0$ (Bill peeks at tails) could be happening.

@@

Kleene Algebra with Tests is an algebraic theory that is defined by
equations and inequalities, which has model classes including
guarded string models, relational models, finite models, and matrix
models.  This paper focuses on defining a a concrete guarded
string algebra, the elements of which are sets of guarded strings, from
an Epik specification.
Definitions and notation mostly follow \cite{kozen2001automata}.
Additional syntax and semantics is included to cover multi-agent
epistemic semantics.  Guarded strings over a finite alphabet ${\sf{P}}$
are like ordinary strings, but with atoms 
alternating with the symbols
from ${\sf{P}}$.  In the algebra described by Figure \ref{epikfigure}, 
is the set of events $\{a_1,a_0,b_1,b_0\}$, and in the elevator example,
$\sf{P}$ is $\{u,d\}$.  In the elevator example, $\sf{B}$ is $\{p,q\}$, and in the coin
example it is $\{h,t\}$.
Assuming a trivially true state formula $\rho$,
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\rho}$
in the elevator example is
$\{\bar q \bar p$, $\bar q p$, $q \bar p$, $q p\}$,
which we write $\{\hat0$, $\hat1$,
$\hat2$, $\hat3\}$.
In the coin example, as we already saw in
\ex{effect},
$\mathcal{A}^\varphi_{\mathsf B}$ is $\{h\bar{t},\bar{h}t\}$,
for which we use the shorthand $\{H,T\}$.
Guarded strings are strings of events, alternating with atoms,
and beginning and ending with atoms.
\ex{gs1} gives the encoding as guarded strings of the worlds in
\ex{elevator1} and \ex{coin1}.
The length of a guarded string
$p$, written $|p|$ is the number of events in $p$.
An atom such as $H$ is a guarded string of length $0$.

\begin{example}   \label{gs1}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}[t]{clc}
{\textit{World}} & {\textit{Guarded string}} & \textit{Length}\\ \hline
$v_1$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2$ & $4$ \\
$v_2$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,u\, \hat 3$ & $3$ \\   
$w_1$ & $H$ & $0$\\
$w_2$ & $H \,a_0\, H$ & $1$ \\  
$w_3$ & $T\,a_0\, T \,b_0\, T$ & $2$\\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

The discussion of \ex{coin2} mentioned building worlds by incrementing
worlds with events.  This is accomplished in guarded string models
with fusion product $\diamond$, a partial operation that combines two
guarded strings, subject to the condition that the atom at the end of
the the first argument is identical to the atom at the start of the
second one.  \ex{fusion1} gives some examples.

\begin{example} \label{fusion1}
\hspace{-2mm}\(\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \; \diamond \; \hat 1 \,u\, \hat 2 = \hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2 \)

\hspace{-2mm}\( \hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \; \diamond \; \hat 2 \,u\, \hat 3 = \textit{undefined} \)

\hspace{-2mm}\(H \; \diamond \; H \,a_1\, H = H \,a_0\, H \)

\hspace{-2mm}\(H \; \diamond \; T \,a_1\, T = \textit{undefined} \)
\end{example}

Rather than individual guarded strings, elements of a guarded string
model for KAT are sets of guarded strings.  In the application, these
elements have the interpretation of propositions (sets of possible
worlds).  In a free guarded string model for KAT, any event can be
adjacent to any atom in a guarded string that is an element of the
underlying set for the algebra.  We instead impose the constraints
coming from the state and effect formulas.  \ex{wf1} defines the
well-formed guarded strings determined by and Epic specification.  (i)
says that each atom is consistent with the state constraint, and (ii)
says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with the effect constraint on $e_i$.\footnote{
An alternative is to define equations such as
$\bar{\phi}=0$ (from the state formula $\phi$)
and $a_1\,{=}\,ha_{1}h$ (from the effect formula $h\mathbin{:}h$ for event $a_1$),
and construct a quotient algebra from the equivalence relation generated
by these equations.  This results in equating sets of guarded strings in the free algebra that
differ by guarded strings that are ill-formed according to the state and
effect formulas.  In the development in the text, we instead use
a set of guarded strings that are well-formed according to the state
and effect formulas as the representative of the equivalence class.
}

\begin{example}   \label{wf1}
Given $\sf{P}$, $\sf{B}$, a
state formula $\varphi$, and an effect formula $\zeta_e$ for each
event $e$ in $\sf{P}$,    
$\alpha_0e_0 ... e_n\alpha_{n+1}$ is well-formed iff

\vspace{1mm}

\hspace{2mm}(i) $\element{\alpha_i}{{\mathcal A}^{\varphi}_{\mathsf{B}}}$ ($0{\leq}i{\leq}n$), and

\hspace{2mm}(ii) $\element{\tuple{\alpha_i,\alpha_{i+1}}}{\sem{\zeta_{ei}}^{\varphi}}$, ($0{\leq} i {\le} n$).
\end{example}

Well-formed guarded strings have the interpretation of worlds in the
application to natural-language semantics.  The set of possible worlds
in the Kripke frame determined by an Epik specification is the set of
well-formed guarded strings.  At this point, we could say in the way
familiar from the type theory of possible worlds semantics that any
set of worlds is a proposition, so that the set of propositions is the
power set of the set of worlds \citep{montague,gallin}.  We will
instead define a more restrictive set of propositions corresponding to the regular sets
of strings. This is deferred to the next section.

In the natural language application, sets of well-formed guarded strings 
have the addional interpretation of event types.
An event-type is something that can ``happen'' is different worlds.
The event of the elevator going up is modeled not as the bare event
symbol $u$ or its unit set $\{u\}$, but as the set of guarded strings
\( \hat 0 \,u\, \hat 1, \hat 1 \,u\, \hat 2,  \hat 2 \,u\, \hat 3\).
The event of the elevator going up in a given world $w$ corresponds
to $w$ being incremented to form $w \diamond x$, where $x$ is an element
of the event type.  \ex{events1} gives the event types for
the examples.

\begin{example} \label{events1}
  \renewcommand{\arraystretch}{1.1}
\(
\begin{array}[t]{l|c}
   &  {\mbox{Event type}}\\ \hline
  u & \{ \hat 0 \,u\, \hat 1, \hat 1 \,u\, \hat 2,  \hat 2 \,u\, \hat 3\}\\
  d & \{ \hat 1 \,d\, \hat 0, \hat 2\,d\,\hat 1, \hat 3 \,d\, \hat 2\}\\
  a_1 & \{H \,a_1\, H\}\\
  a_0 & \{T \,a_0\, T\}\\
  b_1 & \{H \,b_1\, H \}\\
  b_0 & \{T \,b_0\, T \}
\end{array}
\)
  \renewcommand{\arraystretch}{1}
\end{example}

The construction so far defines a set of worlds $W$ from an Epik specification.
Normally and in our examples, $W$ is countably infinite, though some choices of effect
formulas can result in a finite set of worlds.
The next step is to define an alternative relation $R_a$ on worlds for each
agent $a$. This will result in a Kripke frame $\tuple{W,R_a,...}$ in the 
standard sense, consisting of a set of worlds, and a world-alternative relation
for each agent \citep{kripke, hintikka}.
An Epik specification defines an alternative relation on bare events for each
agent $a$, which we notate as $\hat{R}_{a}$.  This should be lifted to a relation
$R_a$ on worlds.  The rough idea is that when a world $w$ is incremented with
an event $e$, in the resulting world $w \diamond e$, epistemic alternatives
for agent $a$ are of the form $w' \diamond e'$, where $w'$ is an alternative to
for $a$ in $w$, and $e'$ is and event-alternative for $a$ (Moore --, Baltag --).\footnote{
 In this it is important that the event-alternative relation for an agent
 is constant across worlds. We anticipate that the definition given here produces
 results equivalent to what is found in literature on event alternatives in
 dynamic epistemic semantics, though we have not verified this.  That literature primarily focuses on mapping
 an epistemic model for a single time and situation to another, and uses general first-order models, rather than
 guarded string models. See ---, and articles in ---.  We picked up the idea from papers by Moss and his colleagues, together
 with Moore () and subsequent literature in situation theory.  This previous literature is motivated by epistemic
 logic and AI planning, rather than natural language semantics as in our application.
} This need to be implemented in a way that takes account of pre- and post-conditions for
events. For this, our approach is to refer the definition of well-formed guarded strings.
\ex{raise1} defines a relation on worlds from a relation on bare events.

\begin{example}  \label{raise1}
Let a set of guarded strings $W$ over events P and primitive tests B, and a relation $\hat{R}$ on P be given.
The corresponding relation $R$ on $W$ holds between a guarded string
\[
\alpha_0e_0 ... e_n\alpha_{n+1} 
\]
in $W$ and a guarded string $q$ in $W$ iff $q$ is of the form

\[
\alpha_0'e_0' ... e_n'\alpha_{n+1}', 
\]

where for $0 \leq n$, $\element{\tuple{e_i,e'_i}}{R}$.

@@

This requires that in the alternative world $q$, each constituent event $e'_i$ is an alternative
to the event $e_i$ in the base world according to $R$.  Compatibilities between events in the alternative
world are enforced by the general definition of worlds. Consider a scenario like the one from Figure 1, with
an additional agent Cem. The base world $T b_0 c_0$ is one where the coin is tails, and first Bob looks at tails,
and them Cem looks at tails. The first event $b_0$ has the alternatives $b_0$ and $b_1$ for Aly,
and the second event $c_0$ has the alternatives $c_0$ and $c_1$ for Aly.
This results in the combinations $b_0c_0$, $b_0c_1$, $b_1c_0$, $b_1c_1$. But these
are filtered by post- and pre-conditions of events in the alternative world, so that
the set of alernatives for Aly in $T b_0 c_0$ is  $\{T b_0 c_0,H b_1 c_1\}$,
with two world-alternatives instead of four.

@@

\begin{example}   \label{Rhathat}
\hspace{-2mm}\(
\hat{R} \triangleq
\hspace{-2mm}\begin{array}[t]{l}
\{ \tuple{x,y} \mid \exists u_1 ... \exists u_n \exists v_1 ... \exists v_n. \\
\hspace{6mm} u_1\mathrel{R}v_1 \wedge ... \wedge u_n\mathrel{R}v_n \wedge\\
\hspace{6mm} x \in \sem{ u_1 ; ... ; u_n }^{\varphi,{\mathcal{E}}}\hspace{1mm} \wedge \\
\hspace{6mm} y \in \sem{ v_1 ; ... ; v_n }^{\varphi,{\mathcal{E}}} \hspace{2mm} \}
\end{array}
\)
\end{example}
